---
title: "Lisa_Wu_HW_3"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(lubridate)
library(arrow)
```

```{r gender}
library(gender)
```

## Load data

Load the following data:
  + applications from `app_data_sample.parquet`
  + edges from `edges_sample.csv`

```{r load-data}
# change to your own path!
data_path <- "/Users/lisawu/Downloads/"
applications <- read_parquet(paste0(data_path,"app_data_sample.parquet"))

applications
```

## Get gender for examiners

We'll get gender based on the first name of the examiner, which is recorded in the field `examiner_name_first`. We'll use library `gender` for that, relying on a modified version of their own [example](https://cran.r-project.org/web/packages/gender/vignettes/predicting-gender.html).

Note that there are over 2 million records in the applications table -- that's because there are many records for each examiner, as many as the number of applications that examiner worked on during this time frame. Our first step therefore is to get all *unique* names in a separate list `examiner_names`. We will then guess gender for each one and will join this table back to the original dataset. So, let's get names without repetition:

```{r gender-1}
library(gender)
#install_genderdata_package() # only run this line the first time you use the package, to get data for it

# get a list of first names without repetitions
examiner_names <- applications %>% 
  distinct(examiner_name_first)

examiner_names
```

Now let's use function `gender()` as shown in the example for the package to attach a gender and probability to each name and put the results into the table `examiner_names_gender`

```{r gender-2}
# get a table of names and gender
examiner_names_gender <- examiner_names %>% 
  do(results = gender(.$examiner_name_first, method = "ssa")) %>% 
  unnest(cols = c(results), keep_empty = TRUE) %>% 
  select(
    examiner_name_first = name,
    gender,
    proportion_female
  )

examiner_names_gender
```

Finally, let's join that table back to our original applications data and discard the temporary tables we have just created to reduce clutter in our environment.

```{r gender-3}
# remove extra colums from the gender table
examiner_names_gender <- examiner_names_gender %>% 
  select(examiner_name_first, gender)

# joining gender back to the dataset
applications <- applications %>% 
  left_join(examiner_names_gender, by = "examiner_name_first")

# cleaning up
rm(examiner_names)
rm(examiner_names_gender)
gc()

```


## Guess the examiner's race

We'll now use package `wru` to estimate likely race of an examiner. Just like with gender, we'll get a list of unique names first, only now we are using surnames.

```{r race-1}
library(wru)

examiner_surnames <- applications %>% 
  select(surname = examiner_name_last) %>% 
  distinct()

examiner_surnames
```
We'll follow the instructions for the package outlined here [https://github.com/kosukeimai/wru](https://github.com/kosukeimai/wru).

```{r race-2}
examiner_race <- predict_race(voter.file = examiner_surnames, surname.only = T) %>% 
  as_tibble()

examiner_race
```

As you can see, we get probabilities across five broad US Census categories: white, black, Hispanic, Asian and other. (Some of you may correctly point out that Hispanic is not a race category in the US Census, but these are the limitations of this package.)

Our final step here is to pick the race category that has the highest probability for each last name and then join the table back to the main applications table. See this example for comparing values across columns: [https://www.tidyverse.org/blog/2020/04/dplyr-1-0-0-rowwise/](https://www.tidyverse.org/blog/2020/04/dplyr-1-0-0-rowwise/). And this one for `case_when()` function: [https://dplyr.tidyverse.org/reference/case_when.html](https://dplyr.tidyverse.org/reference/case_when.html).

```{r race-3}
examiner_race <- examiner_race %>% 
  mutate(max_race_p = pmax(pred.asi, pred.bla, pred.his, pred.oth, pred.whi)) %>% 
  mutate(race = case_when(
    max_race_p == pred.asi ~ "Asian",
    max_race_p == pred.bla ~ "black",
    max_race_p == pred.his ~ "Hispanic",
    max_race_p == pred.oth ~ "other",
    max_race_p == pred.whi ~ "white",
    TRUE ~ NA_character_
  ))

examiner_race
```

Let's join the data back to the applications table.

```{r race-4}
# removing extra columns
examiner_race <- examiner_race %>% 
  select(surname,race)

applications <- applications %>% 
  left_join(examiner_race, by = c("examiner_name_last" = "surname"))

rm(examiner_race)
rm(examiner_surnames)
gc()
```


## Examiner's tenure 

To figure out the timespan for which we observe each examiner in the applications data, let's find the first and the last observed date for each examiner. We'll first get examiner IDs and application dates in a separate table, for ease of manipulation. We'll keep examiner ID (the field `examiner_id`), and earliest and latest dates for each application (`filing_date` and `appl_status_date` respectively). We'll use functions in package `lubridate` to work with date and time values.

```{r tenure-1}
library(lubridate) # to work with dates

examiner_dates <- applications %>% 
  select(examiner_id, filing_date, appl_status_date) 

examiner_dates
```

The dates look inconsistent in terms of formatting. Let's make them consistent. We'll create new variables `start_date` and `end_date`.

```{r tenure-2}
examiner_dates <- examiner_dates %>% 
  mutate(start_date = ymd(filing_date), end_date = as_date(dmy_hms(appl_status_date)))
```

Let's now identify the earliest and the latest date for each examiner and calculate the difference in days, which is their tenure in the organization.

```{r tenure-3}
examiner_dates <- examiner_dates %>% 
  group_by(examiner_id) %>% 
  summarise(
    earliest_date = min(start_date, na.rm = TRUE), 
    latest_date = max(end_date, na.rm = TRUE),
    tenure_days = interval(earliest_date, latest_date) %/% days(1)
    ) %>% 
  filter(year(latest_date)<2018)

examiner_dates
```

Joining back to the applications data.

```{r tenure-4}
applications <- applications %>% 
  left_join(examiner_dates, by = "examiner_id")

rm(examiner_dates)
gc()
```

## Look at examiners demographics

```{r count-examiners}
library(dplyr)
applications %>%
  distinct(examiner_id) %>%
  count()
```

### Compare TCs by gender graphically

```{r compare-tcs-by-gender}
library(dplyr)
library(ggplot2)
applications %>%
  group_by(tc, gender) %>%
  summarise(n = n_distinct(examiner_id)) %>%
  ggplot(aes(x = as.factor(tc), y= n, fill = gender)) +
  geom_col(position = "dodge") +
  xlab ("Technology Center") +
  ylab("Numbers of examiners") +
  ggtitle("Histogram of gender across technology center")
```

#### The graph reveals a higher representation of males in the organization compared to females. However, it is crucial to acknowledge the presence of a significant number of "NAs" in the data, as these figures can greatly impact the interpretation. Examining the ratio between females and males is also noteworthy. For instance, in the case of "tc 1600," the difference between males and females is only slightly pronounced when the NAs are not considered. However, in the case of "tc 2100," the gender gap is noticeably wider.


### Compare TCs by race graphically
```{r compare-tcs-by-race}
applications %>%
  group_by(tc, race) %>%
  summarise(n = n_distinct(examiner_id)) %>%
  ggplot(aes(x = as.factor(tc), y = n, fill = race)) +
  geom_col(position = "dodge") +
  xlab("Technology Center") +
  ylab("Numbers of examiners") + 
  ggtitle("Histogram of race across technology center")
```

#### The data suggests that individuals of white ethnicity constitute the majority across all "tc" categories. However, it is essential to acknowledge the potential bias introduced by name associations. Some individuals may possess names that sound traditionally associated with white ethnicity, despite belonging to other racial backgrounds. Additionally, an intriguing observation emerges: the presence of a higher number of individuals of Asian ethnicity in "tc 2100" and "tc 2400," both of which pertain to computer-related fields.


### Compare TCs by tenure graphically
```{r compare-tcs-by-tenure}
examiners <- applications %>%
  group_by(examiner_id) %>%
  summarise(
    tenure = first(tenure_days), 
    gender = first(gender),
    race = first(race),
    tc = first(tc)
    )

examiners %>%
  ggplot(aes(fill = as.factor(tc), x = tenure)) + 
  geom_histogram(binwidth = 365) +
  ggtitle("Stacked histogram of tenure across technology centres") +
  xlab("Tenure days") +
  ylab("Count of examiners")+
  theme_minimal()

#For this code, why the result for 6000 days above is 600 but not 1500?
#applications %>%
  #group_by(tc, tenure_days) %>%
  #summarise(n = n_distinct(examiner_id)) %>%
  #ggplot(aes(fill = as.factor(tc), x = tenure_days)) + 
  #geom_histogram(binwidth = 365) +
  #ggtitle("Stacked histogram of tenure across technology centres") +
  #xlab("Tenure days") +
  #ylab("Count of examiners")+
  #theme_minimal()
```


#### The presence of a considerable number of employees who have remained with the organization for more than 6000 days indicates a positive aspect for the company. Among the four different "tc" departments there appears to be a relatively similar count of individuals who have surpassed the 6000-day mark in tc 1600, 1700, and 2100. Conversely, tc 2400 exhibits the lowest number in this category. This observation provides a basis for hypothesizing about the turnover rate in each department. Despite tc 1600 and 1700 having fewer employees overall compared to 2100 and 2400, they manage to maintain a similar number of employees who have stayed for more than 6000 days, implying a higher retention rate for these two departments. However, further analysis is necessary to validate this hypothesis.

### Turnover by gender and race 
```{r-turnover}
# Step 1: set 85% sample data and 15% testing data
'''set.seed(123)
examiners <- applications %>%
  group_by(examiner_id) %>%
  summarise(
    tenure = first(tenure_days), 
    gender = first(gender),
    race = first(race),
    tc = first(tc),
    earliest_date = first(earliest_date),
    latest_date = first(latest_date)
    )
main_sample_percentage <- 0.85
examiners_split <- examiners %>%
  mutate(split = ifelse(row_number() <= round(n() * main_sample_percentage), "main", "test"))
main_sample <- examiners_split %>% filter(split == "main")
test_sample <- examiners_split %>% filter(split == "test")
write.csv(main_sample, "main_sample.csv", row.names = FALSE)
write.csv(test_sample, "test_sample.csv", row.names = FALSE)

#Step 2: define turnover: idea is to check the yearly turnover rate and average them. look at who is there the first day of the year and still there the first day of next year, and see how many people held a position at the organization that whole year. then use formula to get a turnover rate. I want to use a while loop to loop 17 years, and then avergae the turnover rate. 

#Create a copy of the dataset to test, so it doesnt mess up my main sample
'''applications <- applications %>%
  mutate(appl_status_date = dmy_hms(appl_status_date))

applications <- applications %>%
  mutate(year = year(appl_status_date))

#filter any data after 2017
applications <- applications %>%
  filter(year <= 2017)

turnover <- applications %>%
  group_by(examiner_id) %>%
  summarize(min_year = min(year), max_year = max(year), tc = first(tc), gender = first(gender), race = first(race)) %>%
  mutate(year_left = if_else(max_year<2017, max_year+1, NA_real_))

turnover_rate2 <- turnover %>%
  group_by(year_left) %>%
  summarize(turnover_count = n()) %>%
  mutate(year = year_left-1)

total_examiners <- applications %>%
  group_by(year) %>%
  summarize(previous_year_count = n_distinct(examiner_id))

turnover_rate2 <- turnover_rate2 %>%
  left_join(total_examiners) %>%
  mutate(turnover_rate = turnover_count/previous_year_count*100) %>%
  select(-year)

```

